---
title: "24_rnaseq_pipeline_RU25"
author: "Anna Zhang"
date: "2026-06-05"
output: html_document
---

# Oyster Data Analysis

## Set Ups

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Load Gene Count Data

```{r}

#################### count - number of reads mapped/gene
######### a measure of gene expression

library(stringr)

# read the gene count files
counts <- read.csv("/work/clh162/Data+/Sophie/2025-2026-team-code-sac165/Data+_team/Sophie/gene_counts/gene_counts.csv", skip = 1, check.names = FALSE, row.names = 1 )

# view/check data format to see if adjustments needed
colnames(counts)[1:10]
head(counts)

# remove featurecounts data to keep only gene data
sample_names <- basename(colnames(counts)[6:ncol(counts)])

# clean sample names
sample_names <- sub("_sorted\\.bam", "", sample_names)
sample_names <- sub("^(.*_Gill).*", "\\1", sample_names)
colnames(counts)[6:ncol(counts)] <- sample_names

# set the column names to just the sample names
count_matrix <- counts[, 6:ncol(counts)]

# save count_matrix
write.table(
  count_matrix,
  file = "/work/clh162/Data+/Anna/2025-2026-team-code-az199/Data+_team/Anna/RU25_gcounts_clean.txt",
  sep = "\t",
  quote = FALSE,
  row.names = TRUE,
  col.names = NA
)
```

### Load and Process Metadata

```{r}

####################### colData - metadata, extra info ex. 
######### location, collection date, rna...etc.
### describes the column of count matrix

# load necessary library
library(readxl)

# read the gene annotation files, skip the first line of "None"
coldata_raw <- read_excel("/work/clh162/OysterRNA24/xRNA for BC Summer 2024.xlsx", sheet = 1, skip = 1)

dim(coldata_raw) #view dimension
```

### Clean and Resize Metadata to Match Gene Counts Data

```{r}
# convert coldata_raw into standard R frame and rename: coldata
coldata <- as.data.frame(coldata_raw)

# set column "Sample_ID" as rownames of coldata
rownames(coldata) <- coldata$Sample_ID
View(coldata) #visualize resulting data

# rheck for matching qualities
all(colnames(count_matrix) == rownames(coldata))

# rearrange rows of coldata to match column names in count matrix
coldata <- coldata[colnames(count_matrix), ]

# Quick quality check
stopifnot(identical(colnames(count_matrix), rownames(coldata)))
all(colnames(count_matrix) == rownames(coldata)) #Matching check

View(coldata) #Visualize
```

### Import and Process Gene Annotations

```{r}

############ gtf is gene annotation
#### contains exons, cds regions, transcripts

# read genes annotations w/ specific attributes
gtf <- read.delim("/work/clh162/OysterRNA24/hisat2_align/Cv_genome_RU_2025_shared/Cvi_RU25.gtf", comment.char= "#", header = FALSE, sep = "\t", quote = "")

#Manually assign geonomic column names
names(gtf) <- c("sequence", "source", "feature", "start", "end", "score", "strand", "frame", "attribute")
table(gtf$feature) #Output counts on CDs, exon, transcripts

# keep transcripts in gtf
transcripts <- gtf[gtf$feature == "transcript", ]
transcripts$attribute <- as.character(transcripts$attribute)

# extract gene_id
transcripts$gene_id <- sub('.*gene_id "([^"]+)".*$', '\\1', transcripts$attribute)
transcripts$gene_id <- as.character(transcripts$gene_id)

# ensure character
rownames(count_matrix) <- as.character(rownames(count_matrix))
transcripts$gene_id <- as.character(transcripts$gene_id)

# intersect
common <- intersect(rownames(count_matrix), transcripts$gene_id)
count_matrix <- count_matrix[common, ]
transcripts <- transcripts[transcripts$gene_id %in% common, ]

# define rowranges
rowranges <- transcripts

# force the identical orders
rowranges <- rowranges[match(rownames(count_matrix), rowranges$gene_id), ]
rownames(rowranges) <- rowranges$gene_id

# final check
stopifnot(identical(rownames(count_matrix), rownames(rowranges)))

```

### Assemble Summarized Experiment

```{r}
stopifnot(
  identical(rownames(count_matrix), transcripts$gene_id) # features
)

# load needed libraries
library(SummarizedExperiment)
library(GenomicRanges)

# convert row annotation to GRanges
genes_row <-GRanges(
  seqnames = rowranges$sequence, 
  ranges = IRanges(rowranges$start, rowranges$end),
  strand = rowranges$strand
)
names(genes_row) <- rownames(rowranges)

# build Summarized Experiment
se <- SummarizedExperiment(
  assays = list(counts = as.matrix(count_matrix)),
  rowRanges = genes_row,
  colData = coldata
)

# quality check
head(assay(se))
head(rowRanges(se))

all(rownames(se) == rownames(counts))
all(colnames(se) == colnames(counts))

colData(se) # access the sample annotations

# save the Summarized Experiment File
dir.create("data", showWarnings = FALSE)
saveRDS(se, "data/ru25_se.rds")
```

### Generate DESeqDataSet

```{r fig.width= 10, fig.height = 8}
#Load required packages
suppressPackageStartupMessages({
    library(DESeq2)
    library(ExploreModelMatrix)})

#Load saved SummarizedExperiment object
se <- readRDS("data/ru25_se.rds")
#Extract month from collection date in data
colData(se)$Month <- format(colData(se)$Date_Collected, "%b")

# Convert metadata variables to factors
colData(se)$Location <- factor(colData(se)$Location)
colData(se)$Month <- factor(
  colData(se)$Month,
  levels = c("Jun", "Jul", "Aug")
)

#View column names to evaluate validity
colnames(colData(se))


#Create DESeqDataSet object
#DESeq for Single Design (Location)
dds_location <- DESeqDataSet(
  se, 
  design = ~ Location
)
dds_location <- DESeq(dds_location)
plotDispEsts(dds_location, main = "Dispersion Estimates: Design ~Location")

#DESeq for Single Design (Month)
dds_month <- DESeqDataSet(
  se, 
  design = ~ Month
)
dds_month <- DESeq(dds_month)
plotDispEsts(dds_month, main = "Dispersion Estimates: Design ~Month")


#DESeq for Combination Design (Month * Location)
dds <- DESeqDataSet(
  se, 
  design = ~ Location * Month
)
dds <- DESeq(dds)
plotDispEsts(dds, main = "Dispersion Estimates: Design ~Location * Month")
```

### Extract, Organize, and Name Data

```{r}
# load necessary packages
library(readr)
library(tibble)

deseq_out_dir <- "/work/clh162/Data+/Anna/2025-2026-team-code-az199/Data+_team/Anna/DESeq2/"
if(!dir.exists(deseq_out_dir)) dir.create(deseq_out_dir, recursive = TRUE)

# Extract CMT Comparisons
res_Jul_vs_Jun_CMT <- results(dds, contrast = c("Month", "Jul", "Jun"), alpha = 0.05)
res_Aug_vs_Jun_CMT <- results(dds, contrast = c("Month", "Aug", "Jun"), alpha = 0.05)
res_Aug_vs_Jul_CMT <- results(dds, contrast = c("Month", "Aug", "Jul"), alpha = 0.05)

# Construct DAF contrasts to be Extracted
deseq_comps <- list(
  "res_DAF_vs_CMT_Jun" = c("Location", "DAF", "CMT"),
  "res_DAF_vs_CMT_Jul" = list(c("Location_DAF_vs_CMT", "LocationDAF.MonthJul")),
  "res_DAF_vs_CMT_Aug" = list(c("Location_DAF_vs_CMT", "LocationDAF.MonthAug")),
  "res_Jul_vs_Jun_DAF" = list(c("Month_Jul_vs_Jun", "LocationDAF.MonthJul")),
  "res_Aug_vs_Jun_DAF" = list(c("Month_Aug_vs_Jun", "LocationDAF.MonthAug")),
  "res_Aug_vs_Jul_DAF" = list(c("Month_Aug_vs_Jun", "LocationDAF.MonthAug"), c("Month_Jul_vs_Jun", "LocationDAF.MonthJul")),
  "res_Jul_vs_Jun_CMT" = res_Jul_vs_Jun_CMT,
  "res_Aug_vs_Jun_CMT" = res_Aug_vs_Jun_CMT,
  "res_Aug_vs_Jul_CMT" = res_Aug_vs_Jul_CMT)

for (item in names(deseq_comps)) {
  message(paste("Processing:", item))
  if (inherits(deseq_comps[[item]], "DESeqResults")) {
    res_df <- as.data.frame(deseq_comps[[item]])
  } else {
    res_obj <- results(dds, contrast = deseq_comps[[item]], alpha = 0.05)
    # --- Assign it to global environment --- #
    assign(item, res_obj, envir = .GlobalEnv)
    res_df <- as.data.frame(res_obj)
  }
}
all_comps <- c(
  #  --- Site-to-Site Comparisons --- #
  "res_DAF_vs_CMT_Jun",
  "res_DAF_vs_CMT_Jul",
  "res_DAF_vs_CMT_Aug",
  
  # ---DAF Comparisons ---#
  "res_Jul_vs_Jun_DAF",
  "res_Aug_vs_Jun_DAF",
  "res_Aug_vs_Jul_DAF",
  
  # --- CMT Comparisons --- #
  "res_Jul_vs_Jun_CMT",
  "res_Aug_vs_Jun_CMT",
  "res_Aug_vs_Jul_CMT"
)

#Convert row names to dedicated column and save as .csv in dir
for (item in all_comps) {
  # --- print saving message to follow workflow ---#
  message(paste("Saving file:", item))
  write_csv(
    as.data.frame(get(item)) |> rownames_to_column(var= "gene_id"),
    paste0(deseq_out_dir, item))}

```

### Remove Unexpressed Genes

```{r}
#Load required packages
suppressPackageStartupMessages({
    library(vsn)
    library(ggplot2)
})

#Number of genes total before filtering
nrow(dds)

# Keep genes with at least 10 counts in at least 3 samples
keep <- rowSums(counts(dds) >= 10) >= 3
sum(keep)
dds_filter <- dds[keep, ]
```

### Adjust Library Size Differences

```{r}
#Library size = the total num of reads assigned to genes/sample

# Total counts per sample
se$libSize <- colSums(assay(se))

#Plot libSize by using R's native pipe "|>"
#to extract colData, turn into reg data frame
#send to ggplot

#Visualize library sizes across all samples
colData(se) |>
  as.data.frame() |>
  ggplot(aes(Sample_ID, 
             y = libSize /1e6, 
             fill = Location)) +   
  geom_col() + 
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, 
                                   hjust = 1, 
                                   vjust = 1))

#Store size factors into  a DESeqDataSet for normalization
dds <- estimateSizeFactors(dds)

#Compare size factors to raw library sizes
ggplot(data.frame(
  libSize = colSums(counts(dds)), 
  sizeFactor = sizeFactors(dds),
  Location = dds$Location), 
       aes(libSize, sizeFactor, 
           color = Location)) +
  geom_point(size = 3) + theme_bw() 
```

### Run DESeq2

```{r}
# Estimate dispersion and fit negative binomial models
dds <- DESeq(dds)

# Extract default DESeq2 results
results <- results (dds)

# Summarize differential expression results
summary(results)

#Give more specific threshold
results0.01 <- results(dds, alpha = 0.01)
summary(results0.01)

#View moodel coefficients
resultsNames(dds)
```

### Transform Data

```{r}
#Examine mean-SD relationships
meanSdPlot(assay(dds), ranks = FALSE)

#Variance stabilizing transformation
vsd <- DESeq2::vst(dds, blind = FALSE)

#Examine mean_SD again after vst
meanSdPlot(assay(vsd), ranks = FALSE)
```

### Create PCAs

```{r}
#Install and load necessary packages
install.packages("ggrepel")
library(ggrepel)            

#1: Create PCA: Color Comparison on Location
pcaData <- plotPCA(vsd,
                   intgroup = c("Location","Month"),
                   returnData = TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))

pcaData$Sample_ID <- rownames(pcaData)
pcaData$Sample_ID <- sub("_Gill$", "", pcaData$Sample_ID)

ggplot(pcaData, aes(x = PC1, y = PC2, label = Sample_ID)) +
  geom_point(aes(color = Location, shape = Month), size = 3) +
  geom_text_repel(aes(label = sub("^.*_([^_]+)_[^_]+$", "\\1", Sample_ID)),size = 3, max.overlaps = Inf, box.padding = 0.3) +
  theme_minimal() +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  coord_fixed() +
  scale_color_manual(values = c(CMT = "red", DAF = "blue"))

#2: Create PCA: Color Comparison on Months

ggplot(pcaData, aes(x = PC1, y = PC2, label = Sample_ID)) +
  geom_point(aes(color = Month, shape = Location), size = 3) +
  geom_text_repel(aes(label = sub("^.*_([^_]+)_[^_]+$", "\\1", Sample_ID)),size = 3, max.overlaps = Inf, box.padding = 0.3) +
  theme_minimal() +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  coord_fixed() +
  scale_color_manual(values = c(Jun = "red", Jul = "orange", Aug = "yellow"))
```

### Volcano Plot

```{r fig.width = 10, fig.height = 8}
library(dplyr)
library(ggplot2)

for (item in all_comps) {
  res_df <- as.data.frame(get(item)) |>
    tibble::rownames_to_column("gene_id") |>
    mutate(
      significant = ifelse(
        !is.na(padj) &
          padj < 0.05 &
          abs(log2FoldChange) >=1,
        "Significant",
        "Not significant"
      )
    )
  p <- ggplot(
    res_df,
    aes(
      x = log2FoldChange,
      y = -log10(padj),
      color = significant
    )
  ) +
    geom_point(alpha = 0.7, size = 2) +
    theme_bw() +
    labs(
      title = paste("Volcano Plot:", sub("^res_", "", item)),
      x = "log2 Fold Change",
      y = "-log10 adjusted p-value",
      color = "Result"
    )

  print(p)
}
```

### Create Heat Maps and Clustering

```{r, fig.width=8, fig.height=6}

# Load necessary packages
library(ComplexHeatmap)
library(matrixStats)
library(dplyr)
library(circlize)

color_rb <- colorRamp2(c(-3,0,3), c("blue", "white", "red"))
#Create vector of comparisons to be looped through
comps <- "Location_DAF_vs_CMT"

#Set factor levels 
vsd$Month <- factor(vsd$Month, levels = c("Jun", "Jul", "Aug"))
vsd$Location <- factor(vsd$Location, levels = c("DAF", "CMT"))
ord <- order(vsd$Location, vsd$Month)

res_df <- as.data.frame(results(dds, name = "Location_DAF_vs_CMT"))
res_df$gene_id <- rownames(res_df)

#Filter out top 50 significantly differentially expressed genes
  topgenes <- res_df|>
    filter(
      !is.na(padj), #remove genes with missing adjusted p-values
      padj < 0.01, #statistically significant
      abs(log2FoldChange)>= 2 #at least 4-fold change 
    ) |>
    arrange(padj) |> # sort from most ot least significant
    slice_head(n = 50) |> #grab the top 50 most significant
    pull(gene_id) #grab the gene ids

  # Z-score each gene
  mat <- assay(vsd)[topgenes,]
  mat <-t(scale(t(mat)))
  mat <- mat[, ord]
  mat[mat< -3] <- -3
  mat[mat> 3]  <- 3
    
  ha <- HeatmapAnnotation(
    Location = vsd$Location[ord],
    Month = vsd$Month[ord],
    col = list(
      Location = c(
        CMT = "darkred",
        DAF = "blue"), 
      Month = c(
        Jun = "gold",
        Jul = "orange",
        Aug = "forestgreen"
      )))
  #Create heatmap sets 
Heatmap(
  mat,
  col = color_rb,
  name = "Gene Expression Level",
  show_row_names = FALSE,
  cluster_columns = FALSE, 
  column_split = data.frame(
    Location = vsd$Location[ord],
    Month = vsd$Month[ord]),
  top_annotation = ha,
  column_title = "Top DE Genes: DAF vs CMT"
    )

# Pairwise Month Heatmaps
loc_loop <- c("CMT", "DAF")
month_loop <- list(
  c("Jun", "Jul"),
  c("Jun", "Aug"),
  c("Jul", "Aug")
)

#Start the pairwise comparison loop
for (loc in loc_loop) {
  dds_heat <- dds[, dds$Location == loc]
  dds_heat$Month <- droplevels(dds_heat$Month)
  design(dds_heat) <- ~ Month
  dds_heat <- DESeq(dds_heat)
  vsd_heat <- vst(dds_heat, blind = FALSE)
  for(obj in month_loop) {
    heat_month1 <- obj[1]
    heat_month2 <- obj[2]
    res_heat <- results(
      dds_heat, 
      contrast = c("Month", heat_month2, heat_month1)
    )
    res_df <- as.data.frame(res_heat)
    res_df$gene_id <- rownames(res_df)
    topgenes <- res_df|>
    filter(
      !is.na(padj), #remove genes with missing adjusted p-values
      padj < 0.01, #statistically significant
      abs(log2FoldChange)>= 2 #at least 4-fold change 
    ) |>
    arrange(padj) |> # sort from most ot least significant
    slice_head(n = 50) |> #grab the top 50 most significant
    pull(gene_id) #grab the gene ids
    
    heat_ord <- order(vsd_heat$Month)
    
    mat <- assay(vsd_heat)[topgenes,]
    mat <-t(scale(t(mat)))
    mat <- mat[, heat_ord]
    mat[mat< -3] <- -3
    mat[mat> 3]  <- 3
    
    heat_ha <- HeatmapAnnotation(
    Month = vsd_heat$Month[heat_ord],
    col = list(
        Month = c(
            Jun = "gold",
            Jul = "orange",
            Aug = "forestgreen")))
    
    #Create HeatMaps
    h_map <- Heatmap(
      mat, 
      col = color_rb, 
      name = "Gene Expression Level",
      show_row_names = FALSE, 
      cluster_columns = FALSE, 
      top_annotation = heat_ha, 
      column_title = paste(
        "Top Genes:",
        loc, 
        heat_month1, 
        "vs.",
        heat_month2
      )
    )
    print(h_map)
  }
}
```

### Pairwise comparison MA Plots

```{r, fig.width=8, fig.height=6}
install.packages("ashr")
 
for (item in all_comps) {
  res_ma <- get(item)
  plotMA(
    res_ma, 
    ylim = c(-4,4),
    main = sub("^res_", "", item)
  )
}
```

### Plot MA with LFC Shrinkage

```{r}
resultsNames(dds)

coef_names <- resultsNames(dds)[-1]

for (item in coef_names) {
  LFC <- lfcShrink(dds, coef = item, type = "apeglm")
  plotMA(
    LFC, 
    main = item, 
    cex = 0.5
  )
  # --- Raw p-value histogram --- #
  hist(
    LFC$pvalue, 
    breaks = 50, 
    col = "grey",
    main = item, 
    xlab = "p-value"
  )
  # --- Adjusted p-value histogram --- #
  hist(
    LFC$padj, 
    breaks = 50, 
    col = "grey",
    main = item, 
    xlab = "Adjusted p-value"
  )
}
```

### Load Environmental Data

```{r}
#Load package needed
library(readr)

#Read the data in excel format
ph <- read_excel("/work/clh162/Data+/Anna/2025-2026-team-code-az199/pH_2024.xlsx")
salinity <- read_excel("/work/clh162/Data+/Anna/2025-2026-team-code-az199/Salinity_2024.xlsx")
dis_ox <- read_excel("/work/clh162/Data+/Anna/2025-2026-team-code-az199/DOsensor_2024.xlsx")
temp <- read_excel("/work/clh162/Data+/Anna/2025-2026-team-code-az199/Salinity_2024.xlsx")

#Standardize data
# Convert all environmental dataset date column to DateTime format
ph$DateTime <- as.Date(ph$DateTime)
salinity$DateTime <- as.Date(salinity$DateTime)
dis_ox$DateTime <- as.Date(dis_ox$DateTime)
temp$DateTime <- as.Date(temp$DateTime)

#Dates when oyster samples were collected for RNA sequencing
collect_dates <- as.Date(c("2024-06-20", "2024-07-22", "2024-08-19"))

# Filter and average environmental matrix for the 7 days leading up to collection date period for all environmental data variables

ph_avg <- bind_rows(lapply(collect_dates, function(d) {
  ph |>
  filter(DateTime >= d - 6,
         DateTime <= d) |>
  summarise(
    CollectionDate = d,
    pH = mean(pH, na.rm = TRUE))
}))
salin_avg <- bind_rows(lapply(collect_dates, function(d) {
  salinity |>
  filter(DateTime >= d - 6,
         DateTime <= d) |>
  summarise(
    CollectionDate = d,
    Salinity = mean(Sal_ppt, na.rm = TRUE))
}))
do_avg <- bind_rows(lapply(collect_dates, function(d) {
  dis_ox |>
  filter(DateTime >= d - 6,
         DateTime <= d) |>
  summarise(
    CollectionDate = d,
    DO = mean(DomgL, na.rm = TRUE))
}))
temp_avg <- bind_rows(lapply(collect_dates, function(d) {
  temp |>
  filter(DateTime >= d - 6,
         DateTime <= d) |>
  summarise(
    CollectionDate = d,
    Temperature = mean(Temp_C, na.rm = TRUE))
}))

#Combine all environmental variables into one master table by collection date
env_master <- ph_avg |>
  left_join(salin_avg, by = "CollectionDate") |>
  left_join(do_avg, by = "CollectionDate") |>
  left_join(temp_avg, by = "CollectionDate")
View(env_master) # Visualize resulting data table
```

### Extract and Convert PCA Dates from Sample ID

```{r}
library(stringr)
pcaData <- pcaData |>
  mutate(CollectionDate = case_when(
    str_detect(Sample_ID, "20Jun24|21Jun24") ~ as.Date("2024-06-20"),
    str_detect(Sample_ID, "22Jul24|23Jul24") ~ as.Date("2024-07-22"),
    str_detect(Sample_ID, "19Aug24|20Aug24") ~ as.Date("2024-08-19"),
    TRUE ~ as.Date(NA)))
pca_mergdat <- left_join(pcaData, env_master, by = "CollectionDate")
```

### Environmental PCAs For Loops

```{r fig.width = 8, fig.height = 6}
# Create list of variables to loop through
env_factors <- c("pH", "Salinity", "DO", "Temperature")


for (item in env_factors) {
  #Extract directional slope coefficient and scale up for visibility purposes
  env_plots <- ggplot(pca_mergdat, aes(x= PC1, y=PC2)) +
    geom_point(aes(color = .data[[item]], shape = Month), size = 3) +
    geom_text_repel(aes(label = sub("^([^_]+_[^_]+)_.*$", "\\1", Sample_ID)),size = 3, max.overlaps = Inf, box.padding = 0.3) +
    scale_color_viridis_c()+
    theme_minimal() +
    labs(
      title = paste("PCA:", item),
      x = paste0("PC1: ", percentVar[1], "% variance"),
      y = paste0("PC2: ", percentVar[2], "% variance"),
      color = item,
      shape = "Month") +
    coord_fixed()
  print(env_plots)
}
```

### Correlations Between Environmental Factors

```{r}
ph_daily <- ph |>
  group_by(DateTime) |>
  summarise(pH = mean(pH, na.rm = TRUE))
salin_daily <- salinity |>
  group_by(DateTime) |>
  summarise(Salinity = mean(Sal_ppt, na.rm = TRUE))
do_daily <- dis_ox |>
  group_by(DateTime) |>
  summarise(DO = mean(DomgL, na.rm = TRUE))
temp_daily <- temp |>
  group_by(DateTime) |>
  summarise(Temperature = mean(Temp_C, na.rm = TRUE))

env_corr <- ph_daily |>
  left_join(salin_daily, by = "DateTime") |>
  left_join(do_daily, by = "DateTime") |>
  left_join(temp_daily, by = "DateTime")
env_corr <-env_corr |>
  filter(DateTime >= as.Date("2024-06-01"),
         DateTime <= as.Date("2024-08-31"))
env_numeric <- env_corr |>
  dplyr::select(pH, Salinity, DO, Temperature)
env_corr_matrix <- cor(env_numeric, 
                       method = "spearman",
                       use = "complete.obs")

#Set color degrations
corr_colors <- colorRamp2(c(-1, 0, 1), c("forestgreen", "white", "orange"))

#Generate Correlation Heatmap for Environmental Factors
env_corr_map <- Heatmap(
  env_corr_matrix, 
  name = "Spearman\nCorrelation (r)",
  col = corr_colors,
  show_row_names = TRUE,
  show_column_names = TRUE, 
  cluster_columns = TRUE,
  cluster_rows = TRUE, 
  column_title = "Correlation Between Environmental Factors",
  cell_fun = function(j, i, x, y, width, height, fill) {
    grid.text(sprintf("%.2f", env_corr_matrix[i,j]), x, y, gp = gpar(fontsize = 10))
  },
  heatmap_legend_param = list(
    at= c(-1, -0.5, 0, 0.5, 1),
    labels = c("-1.0", "-0.5", "0.0", "0.5", "1.0")
  )
)
print(env_corr_map)

#  0.5 correlation between pH & Salinity (interesting), visualize through scatterplot
summary(env_corr)
cor.test(env_corr$pH, env_corr$Salinity)
ggplot(env_corr, aes(Salinity, pH)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal()

```

### WGCNA: Install/load packages

```{r}
# Install/load necessary packages
library(WGCNA)
library(tidyverse)
library(CorLevelPlot)
library(gridExtra)
library(SummarizedExperiment)
library(matrixStats)
allowWGCNAThreads()
```

### WGCNA: Prepare Data

```{r}

#Variance stabilizing transformation
vsd <- DESeq2::vst(dds, blind = FALSE)

# 1. Prepare data
# Need raw count data --> loaded into counts
expr <- assay(vsd)   #expression matrix

# Filter for top 8000 Median Absolute Deviation (MAD) genes to limit modules and eliminate noises
gene_mad <- rowMads(expr)
top8000 <- names(sort(gene_mad, decreasing = TRUE)[1:8000])
expr_filtered <- expr[top8000, ]

# Get metadata
pheno <- as.data.frame(colData(vsd))
feature <- as.data.frame(rowData(vsd))[top8000,  , drop = FALSE]

# Transpose the data
datexpr <- t(expr_filtered)


# Row names and column names check
dim(datexpr)
head(rownames(datexpr))
head(colnames(datexpr))

#Check for missing values
gsg <- goodSamplesGenes(datexpr, verbose = 3)
gsg$allOK #should return TRUE to move on
```

### WGCNA: Cluster Samples

```{r fig.width = 8, fig.height=8}
# Create sample tree
samptree <- hclust(dist(datexpr), method = "average")

# Visualize sample tree
plot(samptree, 
     main = "Sample Clustering to Detect Outliers",
     xlab = "",
     sub = "",
     cex.lab = 1.5,
     cex.axis = 1.5, 
     cex.main = 2)

```

### WGCNA: Adjacency: Pairwise Connection

```{r}
# Nodes are considered "connected" if they have significant pairwise correlation association

#Determine appropriate parameter to calculate adjacency matrix

#Stanadard range recommended by WGCNA
powers <- c(1:10, seq(12, 20, by = 2))

#Calculate scale-free topology statistics
sft <- pickSoftThreshold(datexpr, #sample x genes matrix
                         powerVector = powers, 
                         verbose = 5) #print updated progress
                         
# Visualize through 2 side-by-side plots
#Plot 1: Scale-Free Topology Model Fit
par(mar=c(1, 1, 1, 1))
plot(
  sft$fitIndices[,1],
  -sign(sft$fitIndices[,3]) *
  sft$fitIndices[,2],
  xlab = "Soft threshold (Power)",
  ylab = "Scale-Free Topology Fit Index, signed R^2",
  type = "n",
  main = "Scale Independence"
          )
text(
  sft$fitIndices[,1],
  -sign(sft$fitIndices[,3]) *
    sft$fitIndices[,2],
  labels = powers, 
  col = "red"
)
abline(h= 0.8, col = "navy")

#Plot 2: Mean Connectivity
par(mar=c(1, 1, 1, 1))
plot(
  sft$fitIndices[,1],
  sft$fitIndices[,5],
  xlab = "Soft Threshold (power)",
  ylab = "Mean Connectivity",
  type = "n",
  main = "Mean Connectivity")

text(
  sft$fitIndices[,1],
  sft$fitIndices[,5],
  labels = powers, 
  col = "red")
```

### WGCNA: Call Adjacency Function

```{r}
library(WGCNA) # Load package
softPower <- 3 #pick soft_thresholding power based on Scale Independence graph
```

### WGCNA: Contruct Network and Detect Modules

```{r}
# Key points: 
# blockwiseModules():
# 1. Computers adjacency internally
# 2. Computes TOM internally
# 3. Hierarchically Clusters Genes
# 4. Detect modules
# 5. Merge similar modules
net <- blockwiseModules(
  datexpr, 
  power = softPower, 
  networkType = "signed",
  minModuleSize = 50, 
  maxBlockSize = 10000,
  detectCutHeight = 0.995,
  deepSplit = 2, 
  mergeCutHeight = 0.25,
  numericLabels = TRUE, 
  pamRespectsDendro = FALSE,
  saveTOMs = FALSE, 
  verbose = 3
)
# Test with different parameters, specifically with deepSplit and mergeCutHeight to produce ~12 modules
```

### Convert Numeric Labels to Colors for Visualization

```{r}
table(net$colors)

merged_colors <- labels2colors(net$colors)
unmerged_colors <- labels2colors(net$unmergedColors)

#Print table labeling with colors to numbers
table(merged_colors)
table(unmerged_colors)
```

### Plotting Dendrograms

```{r}
#Plot the dendrograms with labels and given parameters
plotDendroAndColors(
  dendro = net$dendrograms[[1]],
  colors = cbind(unmerged_colors, merged_colors),
  groupLaels = c("Unmerged", "Merged"),
  dendroLabels = FALSE, 
  addGuide = TRUE,
  hang = 0.03, 
  guideHang = 0.05, 
  main = "Single-Block Gene Dendrogram and Module Colors"
)
```

### Module Eigengenes Identification

```{r}
#Create eigengenes list
MElist <- moduleEigengenes(datexpr, colors = net$colors)
MEs <- MElist$eigengenes
head(MEs)

#Save the network
saveRDS(net, "data/WGCNA_network.rds")
```

### WGCNA: Module Merging

```{r}
# Cluster modules based on pairwise eigengenes correlation
# Merge modules that have similar expression profiles

#Calculate eigengene dissimilarity
ME.dissimilarity = 1-cor(MElist$eigengenes, use="complete")

#Clustering eigengenes
METree = hclust(as.dist(ME.dissimilarity), method = "average")
par(mar = c(0,4,2,0)) #setting margin size
par(cex = 0.6); #scaling graphics
plot(METree)
abline(h=0.3, col = "red") #line of 0.25 = correlation of 0.75
```

### WGCNA: Merge Modules

```{r}
#Merge close modules with height cut with correlation coefficient of 0.75 or higher
merge <- mergeCloseModules(datexpr, net$colors, cutHeight = .25)
mergedColors = merge$colors
mergeMEs = merge$newMEs
```

### External Trait Matching

```{r fig.width = 20, fig.height = 12, dpi = 300}
#Load necessary packages
library(dplyr)

#Create trait data collection
traitData <- pheno |>
  mutate(CollectionDate = as.Date(Date_Collected)) |>
           left_join(env_master, by = "CollectionDate")
locationMatrix <- model.matrix(~Location -1, data=traitData)
monthMatrix <- model.matrix(~Month - 1, data = traitData)
traitNumeric <- cbind(
  locationMatrix, 
  monthMatrix, 
  traitData[, c(
    "pH",
    "Salinity",
    "DO",
    "Temperature")]
)

#Correlate module eigengenes with traits
mod_trait_cor <- cor(
  MEs, 
  traitNumeric, 
  use = "pairwise.complete.obs"
)
mod_trait_pval <- corPvalueStudent(
  mod_trait_cor, 
  nSamples = nrow(datexpr)
)
textMatrix <- paste(
  signif(mod_trait_cor, 2),
  "\n(",
  signif(mod_trait_pval, 1),
  ")",
  sep = ""
)
dim(textMatrix) <- dim(mod_trait_cor)

#Plot the heatmap
par(mar = c(8,10,3,3))

labeledHeatmap(
  Matrix = mod_trait_cor, 
  xLabels = colnames(traitNumeric),
  yLabels = names(MEs),
  ySymbols = names(MEs),
  colorLabels = FALSE, 
  colors = blueWhiteRed(50), 
  setStdMargins = FALSE, 
  textMatrix = textMatrix, 
  cex.text = 0.9, 
  zlim = c(-1,1),
  main = "Module-Trait Relationships"
)
table(net$colors)
```

### Matching Genes

```{r}
library(readr)
library(dplyr)
library(stringr)

#Load and strip gene and protein id from 2017 RU (reference mapping table)
ru17_gene_protein <- read_csv("/work/clh162/OysterRNA24/DEG_analysis/GO/Cvi_RU17_ncbi_dataset/data/GCF_002022765.2/Cvi_RU17_gene_proteinID.csv",
                              col_names = c("gene", "protein_id"))

# Load Diamond output results
diamond_cols <- c(
    "RU25_gene", "protein_id", "pident", "length", "mismatch",
  "gapopen", "qstart", "qend", "sstart", "send", "evalue", "bitscore") #defining column names

#contains ortholog prediction data
diamond_out <- read_tsv("/work/clh162/Data+/Anna/2025-2026-team-code-az199/Data+_team/Anna/diamond_results.txt", col_names = diamond_cols)

#Match genes naming
diamond_out <-diamond_out |>
  mutate(RU25_gene = str_replace(RU25_gene, "\\.model\\.", ".TU."))

#Joining genes id and protein
#For each RU25 gene match, find RU17 gene name using protein_id
joined_gene_ids <- diamond_out |>
  left_join(ru17_gene_protein, by = "protein_id") |>
  transmute(RU25_gene, ensembl_gene_id = gene)

#Save output
write_csv(
  joined_gene_ids, 
  "/work/clh162/Data+/Anna/2025-2026-team-code-az199/Data+_team/Anna/joined_gene_ids.csv"
)
```

### GO: Load All Necessary Packages

```{r}
#Installs

my_lib <- "/hpc/home/az199/R/x86_64-pc-linux-gnu-library/4.4"
if (!dir.exists(my_lib)) dir.create(my_lib, recursive = TRUE)
.libPaths(c(my_lib, .libPaths()))

if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager", lib = my_lib)
}

BiocManager::install(version = "3.20", ask = FALSE, lib = my_lib)

pkgs_needed <- c("dplyr", "dbplyr", "DBI", "RSQLite", "biomaRt", "goseq",
                  "clusterProfiler", "DESeq2", "AnnotationDbi", "tidyverse")

pkgs_missing <- pkgs_needed[!sapply(pkgs_needed, requireNamespace, quietly = TRUE, lib.loc = my_lib)]

if (length(pkgs_missing) > 0) {
  BiocManager::install(pkgs_missing, lib = my_lib, update = FALSE, ask = FALSE)
}
```

```{r}
# Load Libraries

library(tidyverse)
library(dplyr)
library(biomaRt)
library(DESeq2)
library(AnnotationDbi)
library(goseq)
library(ggplot2)

packageVersion("dplyr") #encountering issues w/ dyplr and biomaRt
```

```{r}
# Establish connection to Ensembl genes and RU17 genome dataset
mart <- useMart("metazoa_mart",
                dataset = "cvgca002022765v4_eg_gene", # old genome
                host = "https://metazoa.ensembl.org")

# Pull GO ids associated with each genes in RU17
attrs <- getBM(attributes = c("ensembl_gene_id", "transcript_length", "go_id", "namespace_1003"), mart = mart)

# GO annotation table (IMPORTANT: Ensembl IDs only)
gene2cat <- attrs %>%
  filter(go_id != "") %>%
  dplyr::select(ensembl_gene_id, go_id) %>%
  as.data.frame()

# Gene lengths (Ensembl IDs only)
gene_lengths <- attrs %>%
  group_by(ensembl_gene_id) %>%
  summarise(transcript_length = max(transcript_length, na.rm = TRUE)) %>%
  as.data.frame()

length_vector <- gene_lengths$transcript_length
names(length_vector) <- gene_lengths$ensembl_gene_id

```

```{r}
# ---Create empty list to hold finals ---#
joined_results <- list()
all_length_vectors <- list()
all_de_vector <- list()

for (item in all_comps) {
  # --- read individual file ---#
  file_path <-paste0(deseq_out_dir, item)
  res_df <- read_csv(file_path, show_col_types = FALSE)
  
  # ---left-join to add ENSEMBL IDs ---#
  res_joined <- res_df %>%
    left_join(joined_gene_ids, by = c("gene_id" = "RU25_gene")) |>
    filter(!is.na(ensembl_gene_id),
           !is.na(padj))
  res_joined <- res_joined |>
    arrange(padj) |>
    group_by(ensembl_gene_id) |>
    slice(1) |>
    ungroup()
  # --- Build binary DE genes vector to match only genes with lengths --- #
  genes_vector <- as.integer(res_joined$padj < 0.05 & !is.na(res_joined$padj))
  names(genes_vector) <- res_joined$ensembl_gene_id
  
  # --- Restrict vectors to overlap w/ DESeq result universe --- #
  common_genes <- intersect(names(genes_vector),
                            names(length_vector))
  
  sub_DE_genes_vector <- genes_vector[common_genes]
  sub_length_vector <- length_vector[common_genes]
  
  # --- Sanity check validation --- #
  stopifnot(length(sub_DE_genes_vector) == length(sub_length_vector))
  stopifnot(all(names(sub_DE_genes_vector) == names(sub_length_vector)))
  
  # --- Save vectors to empty lists created --- #
  all_length_vectors[[item]] <- sub_length_vector
  all_de_vector[[item]] <- sub_DE_genes_vector}
```

### Running Goseq()

```{r fig.width=10, fig.height=12}
# Set path
goseq_out_dir <- "/work/clh162/Data+/Anna/2025-2026-team-code-az199/Data+_team/Anna/goseq_results/"
if(!dir.exists(goseq_out_dir)) dir.create(goseq_out_dir, recursive = TRUE)

total_GO_terms <- 0

# Create loop for each comparison
for (item in all_comps) {
  # --- 1) Extract DE and length vectors needed for this iteration --- #
  DE_genes_vector <- all_de_vector[[item]]
  length_vector <- all_length_vectors[[item]]
  
  # --- 2) Fit Probability Weighting Function (PWF) to account for gene length biases --- #
  pwf <- nullp(DE_genes_vector, 
               bias.data = length_vector, 
               plot.fit = FALSE) #rejects plots in console
  # --- 3) Run goseq --- #
  GO.wall <- goseq::goseq(pwf = pwf,
                   gene2cat = gene2cat,
                   use_genes_without_cat = FALSE)
  # --- 4) Apply multiple testing correction method of Benjamini-Hochberg (BH) to control the False Discovery Rate (FDR) (aka limit type II errors) --- #
  GO.wall$padj <- p.adjust(GO.wall$over_represented_pvalue, method = "BH")
  # --- 5) Filter for significant terms (e.g., padj < 0.05) --- #
  enriched_GO <-GO.wall |>
    filter(padj < 0.05)
    # zero GO terms are significantly abundant
  # --- 6) Save goseq results to .csv file --- #
  clean_name <- sub("^res_", "", item)
  file_out_path <- file.path(goseq_out_dir, paste0("GO_", clean_name, ".csv")) #define the way path is written out
  write_csv(GO.wall, file = file_out_path)
  
  # Plot top 10 significant GO terms
 p <- GO.wall |>
  mutate(hitsPerc = (numDEInCat * 100 / numInCat) |>
           arrange(over_represented_pvalue) |>
           slice_head(n=10) |>
           mutate(
             term = factor(term, levels = rev(term))),
         log10_p = -log10(pmax(over_represented_pvalue, .Machine$double.xmin))) |>
  ggplot(aes(x = hitsPerc, 
             y = reorder(term, log10_p),
             colour = log10_p, 
             size = numInCat)) +    
  geom_point() +
  expand_limits(x = 0) +
  labs(title = clean_name, x = "Hits (%)", y = "GO term",
       colour = "-log10(p-value)", size = "Count") +
  theme_minimal()
 print(p)
}

```
